|

### 2023-01-18 text/ManuscriptDraft.Rmd
Removed old figures. Updating simulated data figure

---
new_session: FALSE
#title: |
#  **Towards individual-based pollination ecology: Automatic tracking of life histories of individual flowers**
#date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  word_document: default
  pdf_document:
    fig_caption: yes
    keep_tex: yes
    latex_engine: pdflatex
  html_document:
    df_print: paged
fontfamily: mathpazo
fontsize: 12pt
geometry: margin = 1in
header-includes:
- \usepackage{setspace}\doublespacing
- \usepackage[left]{lineno}
- \linenumbers
- \usepackage{dcolumn}
- \usepackage{caption}
- \usepackage{float}
- \usepackage{afterpage}
- \usepackage{siunitx}
- \usepackage{amsmath}
#keywords: Tracking, flowering phenology, Arctic
bibliography: ./library.bib
csl: ./journal-of-ecology.csl
---

@@

Hjalte M. R. Mann^1,2^, Alexandros Iosifidis^2^, Toke T. Høye^1^

@@

1: Department of Ecoscience and Arctic Research Centre, Aarhus University, C.F. Møllers Allé 8, 8000 Aarhus C, Denmark\
2: Department of Electrical and Computer Engineering -- Signal Processing and Machine Learning, Aarhus University, Finlandsgade 22, 8200 Aarhus N, Denmark

@@

@@

 2. Here, we show that image-based monitoring of field plots at very high temporal resolution can return information on flowering phenology at the level of individual flowers. Further, we present a framework for automatically tracking flowers in time-lapse image series, evaluating tracking accuracy, and identifying unreliable tracks.

 3. Our results on individual flower phenology suggest that traditional weekly observations of flower plots are insufficient to capture details in flowering dynamics. We show that the results of our automated method are reliable by comparing with manual tracking of individual floral phenology.

 4. Our method could facilitate fine-grained studies on flower phenology and could be an important tool for estimating the sensitivity of plant-pollinator interactions to climate change.

@@

The Arctic is characterized by short growing seasons and for the individual plant, timing of flowering is of critical importance. Flowering too early increases the risk of frost damage to the flowers, while flowering too late limits the time for seed maturation and dispersal before the end of the growing season [@pardee2019; @ELZINGA2007; @inouye2008; @iler2021]. Further, for insect-pollinated plants, flowering must be synchronous with pollinator activity for successful reproduction. Both flower and insect phenology is partly driven by abiotic cues such as snowmelt and temperature and can thus be sensitive to climate change. Changes in floral phenology as a consequence of climate change may indirectly affect the abundance of flower-dependent insect species [@hoye2013] and these effects can be stronger than the direct effects of climate change [@ogilvie2017]. However, the relation between floral phenology and reproductive success is not well understood. For example, @bolmgren2015 found that for an insect-pollinated plant species, early flowering was associated with increased fruit-set but late flowering was associated with higher seed mass. If we are to understand the intricate dynamics of plant-pollinator interactions, tools that aid efficient and comprehensive data collection are essential.

@@

![Figure 1: Tracking individual flowers through time. A: A time-lapse sequence of five frames showing presence of five individual flowers. Flower a (orange) is present in all frames, while the other flowers are only present part of the sequence and/or have gaps in their presence (e.g. because of occlusion or because they have temporarily disappeared from frame of view). Correct tracking of individual flowers results in information about the distribution of each individual flower in space (B) and time (C).](../figures/flowertrackingOverview.png){width="30%"}

Most multiple object tracking literature focus on tracking of humans [@luo2021; @milan2016], and these methods not necessarily directly transferable to other use-cases. Many tracking algorithms, such as the Simple online and real-time tracking (SORT) [@sort], rely on Kalmann filtering to predict the future location of an object by estimating its velocity from previous frames [@kalmann]. However, Kalman filtering assumes linear motion of the object and does not handle abrupt motion well. Flowers are constrained in movement by their stalk and violate this assumption. Other tracking approaches, such as DeepSORT [@deepsort2017], use tracking-by-detection in which convolutional neural networks (CNNs) are applied to continuously distinguish individuals through time, thus aiding the tracking algorithm. Individual flowers, however, are unlikely to be clearly distinguishable and one flower may look less alike between two frames than two flowers in a single frame, which prevents this approach. Finally, some tracking methods require manual initiation of tracks for a fixed number of objects in the first frame and the algorithm then tracks these objects in the subsequent frames [@hu2012]. These methods do not allow objects to enter or disappear from frame during tracking. In full-season monitoring data, flowers will disappear from the frame when they wilt, in which case their track should be terminated. Oppositely, new flowers should continuously be tracked as they bloom throughout the season. Additionally, a flower tracking method must handle flowers disappearing and appearing at the edges of the frame. Clearly, individual flower tracking is a challenging task.

Here, we present and evaluate a simple multiple object tracking algorithm for automatic tracking of individual flowers in image-series. Our tracking algorithm is based on centroid tracking and assigns individual objects to tracks based on the minimum distance between centroids of bounding boxes and does not rely on prediction of future positions. Further, the method allows both initiation and termination of tracks during tracking. We describe three user-defined parameters incorporated into the algorithm, demonstrate their effect on simulated data, and discuss why they may help optimise tracking of flowers in particular. We apply the algorithm on annotations of real flowers in time-lapse image-series from full-season monitoring of two widespread Arctic flowering plant species *Dryas integrifolia* and *D. octopetala*. For complex scenes with many flowers in close vicinity to each other there is an increased risk of tracking errors regardless of the tracking method being used. For such cases, we suggest a conservative filtering approach based on the distance between flower tracks within the frame in order to identify tracks that are spatially isolated and likely to be tracked with certainty. Overall, we show that information on floral biology can be automatically derived from image-based monitoring of flower plots.

@@

To demonstrate our tracking algorithm and the effect of the three parameters, we used simulated data mimicking flower detections.To produce simulated detections of a flower, we defined a circle with radius rF and center point cF to delimit the boundaries of the flower. The first data point for the flower was then generated as a random point within this circle. Then, remaining data points (i.e. data points for the subsequent frames) were generated iteratively, each point constrained to be within the circle with radius rM and the center point of the previously generated point and within the flower circle.  This approach allowed us to simulate specific scenarios on which to demonstrate our algorithm by controlling the number of flowers generated and the number of data points produced per flower, the boundaries of a flower, and the maximum distance between current and next point (i.e. degree of movement between frames).

We generated four simulated data sets (Fig. 2) to demonstrate each of the three tracking parameters as well as the influence of frame rate on tracking reliability. In simulation A (Fig. 2A), we produced two flowers that are spatially distinct and temporally separated (i.e., one flower appears subsequent of the other). Each flower consists of ten data points and thus the data set spans over 20 frames. In simulation B, we produced four flowers that are spatially separated. Each flower consists of 5 data points, but the data set spans over 10 frames and each flower contains gaps of 1-3 data points at random temporal location. In simulation C, we produced four flowers that are in close proximity to each other and with a large degree of movement between frames. Each flower consist of 500 data points and the data sets spans over 500 frames. Finally, in simulation D, we produced four flowers consisting of a very high number of data points (10.000 per flower) but with only little movement between frames. Here, we used the same center point (cF) for all four flowers to increase overlap. Two pairs of flowers have a high degree of spatial overlap.

@@

![Figure 2: Simulated flower detection data for demonstration of our tracking algorithm and the effect of the three tracking parameters as well as the significance of frame rate for reliable tracking. A: 10 points per flower. rF = 200, rM = 100. B: 10 points per flower. rF = 200, rM = 100 C: 500 points per flower. rF = 200, rM = 200. D: 10,000 points per flower. rF = 100, rM = 1.](../figures/simulatedData.png){width="30%"}

@@

Tracking for simulation A without applying the three tracking parameters resulted in a merging error (i.e. the two objects were tracked as one). Setting the max distance to 200 removed this error and resulted in perfect tracking of the two objects. Tracking for simulation B without applying the tracking parameters returned 8 tracks and 9 mismatches (MOTA = 0.55). Setting max gap to 4 decreased the number of mismatches to 4 (4 tracks, MOTA = 0.8) and including a value of 200 for max distance as for simulation A removes the remaining errors and resulted in perfect tracking. For simulation C, tracking without parameters ignored returned the correct number of tracks (4), but 38 mismatches were made (MOTA = 0.98). Setting the running mean parameter to 20, 40, 60, resulted in 16, 8, and 4 mismatches, respectively, while a running mean value of 80 resulted in perfect tracking. A running mean value of 100 resulted in 4 mismatches. For simulation D, the algorithm tracked the four objects perfectly despite substantial overlap. Basing tracking on every 10th, 20th, and 30th frame resulted in 2, 4, and 10 mismatches, respectively.

>>>>

---
new_session: FALSE
#title: |
#  **Towards individual-based pollination ecology: Automatic tracking of life histories of individual flowers**
#date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  word_document: default
  pdf_document:
    citation_package: natbib
    fig_caption: yes
    keep_tex: yes
    latex_engine: pdflatex
  html_document:
    df_print: paged
fontfamily: mathpazo
fontsize: 12pt
geometry: margin = 1in
header-includes:
- \usepackage{setspace}\doublespacing
- \usepackage[left]{lineno}
- \linenumbers
- \usepackage{dcolumn}
- \usepackage{caption}
- \usepackage{float}
- \usepackage{afterpage}
- \usepackage{siunitx}
- \usepackage{amsmath}
#keywords: Tracking, flowering phenology, Arctic
bibliography: ./library.bib
csl: ./journal-of-ecology.csl
---

@@

Hjalte M. R. Mann^1^, Alexandros Iosifidis^2^, Toke T. Høye^1^

@@

1: Department of Ecoscience and Arctic Research Centre, Aarhus University, C.F. Møllers Allé 8, 8000 Aarhus C, Denmark\
2: Department of Electrical and Computer Engineering, Aarhus University, Finlandsgade 22, 8200 Aarhus N, Denmark

@@

Word count: 
Article type: Research article

@@

 2. Here, we show that image-based monitoring of field plots at very high temporal resolution can return information on flowering phenology at the level of individual flowers. Further, we present a framework for automatic tracking of flowers in time-lapse image series, evaluating tracking accuracy, and identifying unreliable tracks.

 3. Our results on individual floral phenology suggest that traditional weekly observations of flower plots are insufficient to capture details in flowering dynamics. We show that the results of our automated method are reliable by comparing with manual tracking of individual floral phenology.

 4. Our method can facilitate fine-grained studies on floral phenology and could be an important tool for estimating the sensitivity of plant-pollinator interactions to climate change.

@@

The Arctic is characterized by short growing seasons and for the individual plant, timing of flowering is of critical importance. Flowering too early increases the risk of frost damage to flowers, while flowering too late limits the time for seed maturation and dispersal before the end of the growing season [@pardee2019; @ELZINGA2007; @inouye2008; @iler2021]. Further, for insect-pollinated plants, flowering must be synchronous with pollinator activity for successful reproduction. Both flower and insect phenology is partly driven by abiotic cues such as snowmelt and temperature and can thus be sensitive to climate change. Changes in floral phenology as a consequence of climate change may indirectly affect the abundance of flower-dependent insect species [@hoye2013] and these effects can be stronger than the direct effects of climate change [@ogilvie2017]. However, the relation between floral phenology and reproductive success is not well understood. For example, @bolmgren2015 found that for an insect-pollinated plant species, early flowering was associated with increased fruit-set but late flowering was associated with higher seed mass. If we are to understand the intricate dynamics of plant-pollinator interactions, tools that aid efficient and comprehensive data collection are essential.

@@

![Figure 1: Tracking individual flowers through time. A: A time-lapse sequence of five frames showing presence of five individual flowers a (orange), b (purple), c (blue), d (black), and e (green). Flower a is present in all frames, while the other flowers are only present in part of the sequence and/or have gaps in their presence e.g. because of occlusion or because they have temporarily disappeared from frame of view. Correct tracking of individual flowers results in information about the distribution of each individual flower in space (B) and time (C).](../figures/flowertrackingOverview.png){width="30%"}

Most multiple object tracking literature focus on tracking of humans [@luo2021; @milan2016], and these methods are not necessarily directly transferable to other use cases. Many tracking algorithms, such as the Simple online and real-time tracking (SORT) [@sort], rely on Kalmann filtering to predict the future location of an object by estimating its velocity from previous frames [@kalmann]. However, Kalman filtering assumes linear motion of the object and does not handle abrupt motion well. Flowers are constrained in movement by their stalk and violate this assumption. Other tracking approaches, such as DeepSORT [@deepsort2017], use tracking-by-detection in which convolutional neural networks (CNNs) are applied to continuously distinguish individuals through time, thus aiding the tracking algorithm. Individual flowers, however, are unlikely to be clearly distinguishable and one flower may look less alike between two frames than two flowers in a single frame, which prevents this approach. Finally, some tracking methods require manual initiation of tracks for a fixed number of objects in the first frame and the algorithm then tracks these objects in the subsequent frames [@hu2012]. These methods do not allow objects to enter or disappear from frame during tracking. In full-season monitoring data, flowers will disappear from the frame when they wilt, in which case their track should be terminated. Oppositely, new flowers should continuously be tracked as they bloom throughout the season. Additionally, a flower tracking method must handle flowers disappearing and appearing at the edges of the frame. Clearly, individual flower tracking is a challenging task.

Here, we present and evaluate a simple multiple object tracking algorithm for automatic tracking of individual flowers in image-series. Our tracking algorithm is based on centroid tracking and assigns individual objects to tracks based on the minimum distance between centroids of bounding boxes and does not rely on prediction of future positions. Further, the method allows both initiation and termination of tracks during tracking. We describe three user-defined parameters incorporated into the algorithm, demonstrate their effect on simulated data, and discuss why they may help optimise tracking of flowers in particular. We apply the algorithm on annotations of real flowers in time-lapse image-series from full-season monitoring of two widespread Arctic flowering plant species *Dryas integrifolia* and *D. octopetala*. For complex scenes with many flowers in close vicinity to each other, there is an increased risk of tracking errors regardless of the tracking method being used. For such cases, we suggest a conservative filtering approach based on the distance between flower tracks within the frame in order to identify tracks that are sufficiently spatially isolated and likely to be tracked with certainty. Overall, we show that information on floral biology can be automatically derived from image-based monitoring of flower plots.

@@

To demonstrate our tracking algorithm and the effect of the three parameters, we used simulated data mimicking flower detections.To produce simulated detections of a flower, we defined a circle with radius rF and center point cF to delimit the boundaries of the flower. The first data point for the flower was then generated as a random point within this circle. Then, remaining data points (i.e. data points for the subsequent frames) were generated iteratively whith each new point being constrained to be within a circle with a center point of the position of the previously generated point and a radius rM, as well as limited to be within the boundaries of the flower. This approach allowed us to simulate specific scenarios on which to demonstrate our algorithm by controlling the number of flowers generated and the number of data points produced per flower, the boundaries of a flower, and the degree of movement between frames (i.e. the maximum distance between current and next point).

We generated four simulated data sets (Fig. 2) to demonstrate each of the three tracking parameters as well as the influence of frame rate on tracking reliability. In simulation A (Fig. 2A), we produced two flowers that are spatially distinct and temporally separated (i.e., one flower appears subsequent of the other). Each flower consists of ten data points and thus the data set spans over 20 frames. In simulation B, we produced four flowers that are spatially separated. Each flower consists of 5 data points, but the data set spans over 10 frames and each flower contains gaps of 1-3 data points at random temporal location. In simulation C, we produced four flowers that are in close proximity to each other and with a large degree of movement between frames. Each flower consist of 500 data points and the data sets spans over 500 frames. Finally, in simulation D, we produced four flowers consisting of a very high number of data points (10,000 per flower) but with only little movement between frames. Here, we used the same center point (cF) for all four flowers to increase overlap. Two pairs of flowers have a high degree of spatial overlap.

@@

![Figure 2: Simulated flower detection data for demonstration of our tracking algorithm and the effect of the three tracking parameters as well as the significance of frame rate for reliable tracking. A: 10 points per flower. rF = 200, rM = 100. Arrow shows the last point of flower a and the first point of flower b. B: 10 points per flower. rF = 200, rM = 100 C: 500 points per flower. rF = 200, rM = 200. D: 10,000 points per flower. rF = 100, rM = 1.](../figures/simulatedData.png){width="30%"}

@@

Tracking for simulation A without applying the three tracking parameters resulted in a merging error (i.e. the two objects were tracked as one). Setting the max distance to 200 removed this error and resulted in perfect tracking of the two objects. Tracking for simulation B without applying the tracking parameters returned 8 tracks and 9 mismatches (MOTA = 0.55). Setting max gap to 4 decreased the number of mismatches to 4 (4 tracks, MOTA = 0.8) and including a value of 200 for max distance as for simulation A removes the remaining errors and resulted in perfect tracking. For simulation C, tracking without parameters ignored returned the correct number of tracks (4), but 38 mismatches were made (MOTA = 0.98). Setting the running mean parameter to 20, 40, 60, resulted in 16, 8, and 4 mismatches, respectively, while a running mean value of 80 resulted in perfect tracking. A running mean value of 100 resulted in 4 mismatches. For simulation D, the algorithm tracked the four objects perfectly despite substantial overlap, while basing tracking on every 10th, 20th, and 30th frame resulted in 2, 4, and 10 mismatches, respectively.
